\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage[margin=1in]{geometry}

\title{Entropy-Gated Local Learning: Near-Backprop Accuracy Without Backpropagation}
\author{Fabien}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We introduce Entropy-Gated Learning (EG), a local learning algorithm where each neuron
controls its own plasticity based on its output entropy. Neurons with high entropy
(uncertain) learn rapidly; neurons with low entropy (confident) stabilize. Combined with
reconstruction error, decorrelation, and sparsity signals -- all computed locally -- EG
achieves 97.46\% accuracy on MNIST, within 0.58\% of backpropagation (98.04\%), using
zero gradient propagation through feature layers.
\end{abstract}

\section{Introduction}
TODO: Problem statement -- backprop is non-local, biologically implausible, memory-intensive.
Motivation -- local learning rules that scale.

\section{Related Work}
TODO: Forward-Forward (Hinton 2022), Predictive Coding, Equilibrium Propagation,
Hebbian learning, Target Propagation. 16 approaches tested, most fail.

\section{Method}

The core update rule for neuron $i$:
\begin{equation}
\Delta W_i = \sigma\bigl(k(H_i - \theta)\bigr) \cdot \bigl(\alpha_1 R_i + \alpha_2 D_i + \alpha_3 S_i\bigr)
\end{equation}
where:
\begin{align}
H_i &= -\frac{1}{B}\sum_{b=1}^{B}\bigl[y_i^{(b)}\log y_i^{(b)} + (1-y_i^{(b)})\log(1-y_i^{(b)})\bigr] \\
R_i &= y^\top(x - yW)/B \quad \text{(reconstruction)} \\
D_i &= -(C_{\text{offdiag}} \cdot W) \quad \text{(decorrelation)} \\
S_i &= -(\bar{y}_i - \tau)W \quad \text{(sparsity)}
\end{align}

\section{Experiments}
TODO: MNIST results table (V1 through V4), CIFAR-10 results, comparison to backprop.

\section{Analysis}
TODO: Why entropy gating works (implicit curriculum), why 2 layers optimal,
why data augmentation hurts, stability analysis.

\section{Conclusion}
TODO: Summary, 0.58\% gap on MNIST, open challenge on CIFAR-10, future directions.

\bibliographystyle{plain}
\bibliography{refs}

\end{document}
