\documentclass[11pt,a4paper]{article}
\usepackage{amsmath, amssymb, graphicx, hyperref}
\usepackage{geometry}
\geometry{margin=1in}

\title{\textbf{Latent Vacuum Stationarity as a Planck-Scale Boundary Condition}\\
\large Predicting Gravitational Anomalous Dimensions from the Standard Model}
\author{Fabien}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
The Standard Model (SM) of particle physics exhibits a remarkable "near-criticality" around the Planck scale ($M_{Pl} \sim 1.22 \times 10^{19}$ GeV), where both the Higgs quartic coupling $\lambda$ and its $\beta$-function closely approach zero. Within the Latent Vacuum Stationarity (LVS) framework, observable reality is postulated to correspond strictly to the stationary fixed points of the quantum vacuum. In this paper, we demonstrate that interpreting LVS as a global principle that minimizes the Renormalization Group (RG) flow length is falsified by the empirical strong coupling $\alpha_s(M_Z)$. However, interpreting LVS as an exact boundary condition at the Planck scale fundamentally shifts the predictive paradigm. By enforcing strict stationarity ($\beta_i = 0$) at $M_{Pl}$ within the Asymptotic Safety scenario, we algebraically invert the RG equations. Instead of predicting SM parameters from uncertain Quantum Gravity (QG) models, we use high-precision empirical values of the SM to predict the necessary gravitational anomalous dimensions: $f_g \approx 0.0105$ and $f_y \approx 0.0078$. This provides a falsifiable, top-down benchmark for functional RG and lattice quantum gravity calculations.
\end{abstract}

\section{Introduction}
The coincidence of $\lambda \to 0$ and $\beta_\lambda \to 0$ near the Planck scale is often regarded as an accidental cancellation in the Standard Model. LVS proposes that stability is the criterion of physical existence. Thus, this near-criticality is not an accident but the shadow of an exact boundary condition.

\section{Falsification of Global Stationarity}
We first test a naive formulation of LVS: that the parameters of the SM are chosen to globally minimize the variation of the RG flow from the electroweak scale $M_Z$ to $M_{Pl}$. Minimizing the relative length of the flow with respect to $\alpha_s(M_Z)$ yields an optimum of $\alpha_s \approx 0.056$. The measured value is $\alpha_s = 0.1179 \pm 0.0010$. This conclusively falsifies the global minimization hypothesis.

\section{LVS as a Planck-Scale Boundary Condition}
We localize LVS at the Planck scale, demanding exact stationarity. In the absence of gravity, $\beta_{g3} \neq 0$ and $\beta_{yt} \neq 0$ at $M_{Pl}$. In the Asymptotic Safety scenario, quantum gravity introduces anomalous dimensions $f_g$ and $f_y$:
\begin{align}
\beta_{g3} &= \beta_{g3}^{SM} - f_g g_3 = 0 \\
\beta_{yt} &= \beta_{yt}^{SM} - f_y y_t = 0
\end{align}

\section{Inverting the Prediction}
Using the precise boundary values at $M_{Pl}$ derived by Buttazzo et al. (2013)---$y_t = 0.3825$ and $g_3 = 0.4873$---we solve the stationarity conditions algebraically.
For the strong coupling:
\begin{equation}
f_g = \frac{\beta_{g3}^{SM}}{g_3} = b_3 g_3^2 \approx 0.0105
\end{equation}
For the top Yukawa:
\begin{equation}
f_y = \frac{\beta_{yt}^{SM}}{y_t} = c_2 g_3^2 - c_1 y_t^2 \approx 0.0078
\end{equation}

\section{Conclusion and Falsifiability}
By demanding that the vacuum is exactly stationary at the Planck scale, LVS transforms the empirical masses of the Standard Model into direct predictions for Quantum Gravity. If future functional RG calculations of the Reuter fixed point yield anomalous dimensions incompatible with $f_g \approx 0.0105$ and $f_y \approx 0.0078$, the LVS boundary condition is falsified.

\end{document}
