\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage[margin=1in]{geometry}

\title{Fixed-Point Substrate: Intelligence from Equilibrium in a Learnable Medium}
\author{Fabien}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We propose the Fixed-Point Substrate (FPS), a computational paradigm where intelligence
emerges from the equilibrium of a learnable medium rather than from signal propagation
through layers. Input perturbs a medium with local physical properties (conductivity,
coupling, bias), which collapses to a fixed point $Z^* = f(Z^*)$. The stable
configuration IS the computation. Inspired by LVS theory ("It from Fix"),
FPS achieves 96.44\% on MNIST with zero backpropagation, constant memory,
and convergence in only 7 iterations via Anderson acceleration.
\end{abstract}

\section{Introduction}
TODO: The void problem -- neural networks operate in empty space with no underlying medium.
LVS theory -- reality emerges from fixed points. The substrate hypothesis.

\section{Related Work}
TODO: Deep Equilibrium Models (DEQ), Neural ODEs, implicit layers, Hopfield networks.

\section{Method}

The medium finds its equilibrium via fixed-point iteration:
\begin{equation}
Z^{(k+1)} = \tanh\bigl(\text{MultiLap}(Z^{(k)}) + \text{Mix}(Z^{(k)}) + \alpha X + \beta\bigr)
\end{equation}
where MultiLap uses multi-scale Laplacian operators ($3\times3$, $5\times5$, $7\times7$)
and Mix is channel mixing. Convergence accelerated by Anderson acceleration.

\section{Experiments}
TODO: MNIST 96.44\%, Fashion-MNIST 77.15\%, CIFAR-10 42.24\%. Version evolution table.

\section{Analysis}
TODO: Why fixed-point beats PDE temporal. Readout matters more than medium quality.
Learnable spatial filters. Anderson acceleration analysis.

\section{Conclusion}
TODO: Summary, connection to LVS theory, multi-resolution as next step.

\bibliographystyle{plain}
\bibliography{refs}

\end{document}
